---
title: "Analysis of interaction between specific substance use and poly substance use"
author: "Heather Kates"
date: "2024-09-20"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r,echo=FALSE}
#Load required libraries
library(dplyr)
library(logistf)
library(ggplot2)
library(tidyr)
select <- dplyr::select
```

### The original dataset is used with no subsetting or re-coding

This data is cols AU:BB from sheet "Master Sheet" in https://www.dropbox.com/scl/fi/j37pkkg8624byolqf3aim/MPH-Dataset-Stata.xlsx?rlkey=hdsylsmkursvzazd7a8yc9tk7&st=n00hxd0w&dl=0

```{r}
# Read in the data
data <- read.csv("original_data/Hale2023_originalData.csv")
colnames(data) <- gsub("\\.","_",colnames(data))
colnames(data) <- gsub("__","_",colnames(data))
data <- na.omit(data)
# Convert binary variables (substance use) to factors
data_factor <- data %>%
  mutate(across(c(AVDRM, Marijuana, Amphetamine, Cocaine, Sedatives, Opiates, Phencyclidine, Poly_Substance),
                ~ factor(.x, levels = c(0, 1), labels = c(0, 1))))
#write.csv(data_factor,"data/TableS1.csv",row.names = FALSE)
```

Read in demographic data and merge with data

```{r}
demo_data <- read.csv("original_data/Demographic_data.csv")
data <- merge(data,demo_data,by="PATID")
```

### Build and test all possible models and compare model fit values (AIC) to see which model best fits the data and and whether the factors in that model are significant predictors of AVDRM

```{r}
# Create a list to store the models and their AIC/BIC values along with significant factors and p-values
model_comparisons <- list()

# Define the base formula for the model without interactions
base_formula <- "AVDRM ~ Marijuana + Amphetamine + Cocaine + Sedatives + Opiates + Poly_Substance"

# Define interaction terms for each substance with Poly_Substance
interaction_terms <- c("Marijuana:Poly_Substance", "Amphetamine : Poly_Substance", "Cocaine : Poly_Substance", 
                       "Sedatives : Poly_Substance", "Opiates : Poly_Substance")

# Function to fit the model and store AIC, BIC, significant factors, and their p-values
fit_and_store_model <- function(formula_str, data_factor) {
  model <- glm(as.formula(formula_str), family = binomial(link = "logit"), data = data_factor)
  aic <- AIC(model)
  bic <- BIC(model)
  
  # Extract p-values and significant factors
  summary_model <- summary(model)
  p_values <- coef(summary_model)[, "Pr(>|z|)"]
  
  # Remove the intercept from significant factors
  significant_factors <- names(p_values)[p_values < 0.05 & names(p_values) != "(Intercept)"]
  significant_pvals <- p_values[p_values < 0.05 & names(p_values) != "(Intercept)"]
  
  return(list(formula = formula_str, AIC = aic, BIC = bic, significant_factors = significant_factors, 
              p_values = significant_pvals, model = model))
}

# Create a list to store the models and their AIC/BIC values along with significant factors and p-values
model_comparisons <- list()

# Fit the base model (without any interactions)
base_model <- fit_and_store_model(base_formula, data_factor)
model_comparisons[["Base Model (No Interactions)"]] <- base_model

# Fit models with each individual interaction term
for (interaction in interaction_terms) {
  formula_with_interaction <- paste(base_formula, "+", interaction)
  model_name <- paste("Model with", interaction)
  model_comparisons[[model_name]] <- fit_and_store_model(formula_with_interaction, data_factor)
}

# Fit a model with all interactions
all_interactions_formula <- paste(base_formula, "+", paste(interaction_terms, collapse = " + "))
model_comparisons[["All Interactions"]] <- fit_and_store_model(all_interactions_formula, data_factor)

# Create a data frame to store AIC, BIC, significant factors, and p-values for all models
model_results <- data.frame(
  Model = names(model_comparisons),
  AIC = sapply(model_comparisons, function(x) x$AIC),
  BIC = sapply(model_comparisons, function(x) x$BIC),
  Significant_Factors = sapply(model_comparisons, function(x) ifelse(length(x$significant_factors) == 0, 
                                                                     "None", 
                                                                     paste(x$significant_factors, collapse = ", "))),
  P_Values = sapply(model_comparisons, function(x) ifelse(length(x$p_values) == 0, 
                                                          "None", 
                                                          paste(round(x$p_values, 5), collapse = ", ")))
)
```

```{r,echo=FALSE}
# sort by AIC to see the best models
model_results_sorted <- model_results[order(model_results$AIC), ]

# Change "Significant_Factors" and "P_Values" to character strings
model_results_sorted$Significant_Factors <- as.character(model_results_sorted$Significant_Factors)
model_results_sorted$P_Values <- as.character(model_results_sorted$P_Values)
```

```{r,echo=FALSE}
library(DT)
# Adjust column widths
datatable(model_results_sorted, 
          rownames = FALSE, 
          colnames = c("Model", "AIC", "BIC", "Significant Factors", "P-Values"),
          options = list(
            autoWidth = TRUE,
            columnDefs = list(
              list(width = '200px', targets = 0),  # Adjust width for 'Model'
              list(width = '100px', targets = c(1, 2)),  # AIC, BIC widths
              list(width = '150px', targets = 3),  # Significant Factors width
              list(width = '100px', targets = 4)  # P-Values width
            ),
            scrollX = TRUE
          )
)
#write.csv(model_results_sorted,"results/TableS2.csv")
```

### Get the odds ratio for the best model

```{r}
best_model <- glm(AVDRM ~ Marijuana * Poly_Substance + Amphetamine + Cocaine + Sedatives + Opiates,
                  data = data_factor, family = binomial())

# Calculate odds ratios and confidence intervals
best_model_coefs <- coef(best_model)
best_model_or <- exp(best_model_coefs)  # Odds ratios
conf_int <- exp(confint(best_model))    # Confidence intervals

# Extract p-values from the model summary
best_model_summary <- summary(best_model)
p_values <- coef(best_model_summary)[, "Pr(>|z|)"]  # Extract p-values

# Combine odds ratios, confidence intervals, and p-values into a data frame
odds_ratio_df <- data.frame(
  Predictor = names(best_model_or),
  Odds_Ratio = round(best_model_or, 3),
  CI_95_Lower = round(conf_int[, 1], 3),
  CI_95_Upper = round(conf_int[, 2], 3),
  P_Value = round(p_values, 5)
)

# View the table with odds ratios, confidence intervals, and p-values
print(odds_ratio_df)

# Save the table to a CSV file
write.csv(odds_ratio_df, "results/Table2.csv", row.names = FALSE)
```
