---
title: "Analysis of interaction between specific substance use and poly substance use"
author: "Heather Kates"
date: "2024-09-20"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r,echo=FALSE}
#Load required libraries
library(dplyr)
library(logistf)
library(ggplot2)
library(tidyr)

# Load the DT package
library(DT)
select <- dplyr::select

#Function
# Load necessary library
library(dplyr)

# Define the function
logistf_tablize <- function(model_summary) {
  # Create a data frame for reporting
  report_df <- data.frame(
    Variable = names(model_summary$coefficients)[-1],  # Remove the intercept
    P_Value = model_summary$prob[-1],  # P-values (excluding intercept)
    Odds_Ratio = exp(model_summary$coefficients[-1]),  # Odds Ratios
    CI_Lower = exp(model_summary$ci.lower[-1]),  # Lower bound of 95% CI
    CI_Upper = exp(model_summary$ci.upper[-1])   # Upper bound of 95% CI
  )
  
  # Round values for clarity
  report_df <- report_df %>%
    mutate(
      P_Value = round(P_Value, 4),
      Odds_Ratio = round(Odds_Ratio, 2),
      CI_Lower = round(CI_Lower, 2),
      CI_Upper = round(CI_Upper, 2)
    )
  
  return(report_df)
}
```

# Methods

## Participants

We obtained data through the National NeuroAIDS Tissue Consortium (NNTC) Data Coordinating Center, which included data from the NNTC and Central Nervous System HIV Anti-Retroviral Therapy Effects Research (CHARTER) between January 2005 and September 2016. The criteria for participant inclusion and data collection for individual cohorts have been previously described in published manuscripts (17-19). From these cohorts, Mukerji et al. (4) then compiled participants into a cohort to analyze for resistance mutations (4). Eligibility criteria for the current study included: HIV+ status, at least one paired plasma and cerebrospinal fluid viral load measurement, at least one urinalysis of substance use, and at least 18 years of age. 

## Outcome of Interest

The outcome of interest was the presence of an anti-retroviral drug-resistance mutation (ARV-DRM), which was scored as positive if participants had a mutation in either the cerebrospinal fluid or plasma for either protease inhibitors (PI) or nucleoside reverse transcriptase inhibitors/ non-nucleoside reverse transcriptase inhibitors (NRTI/ NNRTI) as determined by lab tested plasma and cerebrospinal fluid for for HIV ARV-DRM and viral levels.

## Predictors of Interest

**Substance use**. The substances included in our analyses were: cannabis, amphetamine, cocaine, sedatives, and opiates. Substance use status was determined by aggregated urinalysis data that were collected multiple times during the duration of a study participant’s enrollment. The urinalysis data were aggregated such that if a participant screened positive for a substance during their duration in the study, they were scored positive for that substance. If they never screened positive for that substance, they were considered negative. This was performed to reduce potential bias due to inconsistent numbers of visits across participants. Some substance categories in the urinalysis were combined due to the similarity of the substances (e.g. amphetamines and methamphetamine; opioids, oxycodone, buprenorphine, and methadone) with respect to their actions on the central nervous system (20). Participants that screened positive for two or more substances were scored as positive for polysubstance use.

**Other Variables Considered**
We captured and considered the relationship between non-substance use factors on substance use or on the outcome variable. These included sociodemographic characteristics (age at survey enrollment, race/ethnicity, and sex) and study-related characteristics (length of Infection).

This data is cols B:O and AU:BB from sheet "Master Sheet" in https://www.dropbox.com/scl/fi/j37pkkg8624byolqf3aim/MPH-Dataset-Stata.xlsx?rlkey=hdsylsmkursvzazd7a8yc9tk7&st=n00hxd0w&dl=0

```{r}
data <- read.csv("original_data/Hale2023_MasterData.csv")
#omit participants without urinalysis results
data <- data %>% filter(!is.na(Cannabis))
data <- data %>% select_if(~!all(is.na(.)))
```

```{r}
# Create a categorical "Race" variable based on the binary race columns
data$Race <- ifelse(data$MultiRacial == 1, "MultiRacial",
                                ifelse(data$Asian == 1, "Asian",
                                ifelse(data$Black == 1, "Black",
                                ifelse(data$Hispanic == 1, "Hispanic",
                                ifelse(data$Native_American == 1, "Native_American",
                                ifelse(data$Native_Hawaiin == 1, "Native_Hawaiin",
                                ifelse(data$White == 1, "White", "Unknown")))))))

# Convert "Race" to a factor
data$Race <- as.factor(data$Race)
```

```{r}
#Rename gender to sex
data <- data %>% dplyr::rename("Sex"="Gender")
# Convert binary variables to factors explicitly
data <- data %>%
  mutate(across(
    c(ARV.DRM, Cannabis, Amphetamine, Cocaine, Sedatives, Opiates, PolySubstance),
    ~ factor(.x, levels = c(0, 1), labels = c("Negative", "Positive"))
  ))
data$Sex <- factor(data$Sex, levels = c(1, 2), labels = c("Female", "Male"))
#write.csv(data_factor,"data/TableS1.csv",row.names = FALSE)
```

#Create Table 1

```{r}
library(dplyr)

# Calculate total sample size
total_n <- nrow(data)

# Calculate total counts for each ARV-DRM group
total_positive <- nrow(filter(data, ARV.DRM == "Positive"))
total_negative <- nrow(filter(data, ARV.DRM == "Negative"))

# Helper function to calculate n and percent for a given filter
calc_n_and_percent <- function(subset_data, total) {
  n <- nrow(subset_data)
  percent <- (n / total) * 100
  return(list(n = n, percent = percent))
}

# Create a list to store results for each characteristic
results <- list()

# Calculate n and percent for Sex (Total, ARV-DRM Positive, ARV-DRM Negative)
results$Sex_Female_Total <- calc_n_and_percent(filter(data, Sex == "Female"), total_n)
results$Sex_Female_Pos <- calc_n_and_percent(filter(data, Sex == "Female" & ARV.DRM == "Positive"), total_positive)
results$Sex_Female_Neg <- calc_n_and_percent(filter(data, Sex == "Female" & ARV.DRM == "Negative"), total_negative)
results$Sex_Male_Total <- calc_n_and_percent(filter(data, Sex == "Male"), total_n)
results$Sex_Male_Pos <- calc_n_and_percent(filter(data, Sex == "Male" & ARV.DRM == "Positive"), total_positive)
results$Sex_Male_Neg <- calc_n_and_percent(filter(data, Sex == "Male" & ARV.DRM == "Negative"), total_negative)

# Calculate n and percent for Race/Ethnicity (Total, ARV-DRM Positive, ARV-DRM Negative)
results$Race_White_Total <- calc_n_and_percent(filter(data, Race == "White"), total_n)
results$Race_White_Pos <- calc_n_and_percent(filter(data, Race == "White" & ARV.DRM == "Positive"), total_positive)
results$Race_White_Neg <- calc_n_and_percent(filter(data, Race == "White" & ARV.DRM == "Negative"), total_negative)
results$Race_Asian_Total <- calc_n_and_percent(filter(data, Race == "Asian"), total_n)
results$Race_Asian_Pos <- calc_n_and_percent(filter(data, Race == "Asian" & ARV.DRM == "Positive"), total_positive)
results$Race_Asian_Neg <- calc_n_and_percent(filter(data, Race == "Asian" & ARV.DRM == "Negative"), total_negative)
results$Race_Black_Total <- calc_n_and_percent(filter(data, Race == "Black"), total_n)
results$Race_Black_Pos <- calc_n_and_percent(filter(data, Race == "Black" & ARV.DRM == "Positive"), total_positive)
results$Race_Black_Neg <- calc_n_and_percent(filter(data, Race == "Black" & ARV.DRM == "Negative"), total_negative)
results$Race_Hispanic_Total <- calc_n_and_percent(filter(data, Race == "Hispanic"), total_n)
results$Race_Hispanic_Pos <- calc_n_and_percent(filter(data, Race == "Hispanic" & ARV.DRM == "Positive"), total_positive)
results$Race_Hispanic_Neg <- calc_n_and_percent(filter(data, Race == "Hispanic" & ARV.DRM == "Negative"), total_negative)
results$Race_MultiRacial_Total <- calc_n_and_percent(filter(data, Race == "MultiRacial"), total_n)
results$Race_MultiRacial_Pos <- calc_n_and_percent(filter(data, Race == "MultiRacial" & ARV.DRM == "Positive"), total_positive)
results$Race_MultiRacial_Neg <- calc_n_and_percent(filter(data, Race == "MultiRacial" & ARV.DRM == "Negative"), total_negative)
results$Race_NativeAmerican_Total <- calc_n_and_percent(filter(data, Race == "Native_American"), total_n)
results$Race_NativeAmerican_Pos <- calc_n_and_percent(filter(data, Race == "Native_American" & ARV.DRM == "Positive"), total_positive)
results$Race_NativeAmerican_Neg <- calc_n_and_percent(filter(data, Race == "Native_American" & ARV.DRM == "Negative"), total_negative)
results$Race_Unknown_Total <- calc_n_and_percent(filter(data, Race == "Unknown"), total_n)
results$Race_Unknown_Pos <- calc_n_and_percent(filter(data, Race == "Unknown" & ARV.DRM == "Positive"), total_positive)
results$Race_Unknown_Neg <- calc_n_and_percent(filter(data, Race == "Unknown" & ARV.DRM == "Negative"), total_negative)

# Calculate n and percent for Substance Use (Total, ARV-DRM Positive, ARV-DRM Negative)
results$Substance_Cannabis_Total <- calc_n_and_percent(filter(data, Cannabis == "Positive"), total_n)
results$Substance_Cannabis_Pos <- calc_n_and_percent(filter(data, Cannabis == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Substance_Cannabis_Neg <- calc_n_and_percent(filter(data, Cannabis == "Positive" & ARV.DRM == "Negative"), total_negative)
results$Substance_Cocaine_Total <- calc_n_and_percent(filter(data, Cocaine == "Positive"), total_n)
results$Substance_Cocaine_Pos <- calc_n_and_percent(filter(data, Cocaine == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Substance_Cocaine_Neg <- calc_n_and_percent(filter(data, Cocaine == "Positive" & ARV.DRM == "Negative"), total_negative)
results$Substance_Amphetamines_Total <- calc_n_and_percent(filter(data, Amphetamine == "Positive"), total_n)
results$Substance_Amphetamines_Pos <- calc_n_and_percent(filter(data, Amphetamine == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Substance_Amphetamines_Neg <- calc_n_and_percent(filter(data, Amphetamine == "Positive" & ARV.DRM == "Negative"), total_negative)
results$Substance_Opiates_Total <- calc_n_and_percent(filter(data, Opiates == "Positive"), total_n)
results$Substance_Opiates_Pos <- calc_n_and_percent(filter(data, Opiates == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Substance_Opiates_Neg <- calc_n_and_percent(filter(data, Opiates == "Positive" & ARV.DRM == "Negative"), total_negative)
results$Substance_Sedatives_Total <- calc_n_and_percent(filter(data, Sedatives == "Positive"), total_n)
results$Substance_Sedatives_Pos <- calc_n_and_percent(filter(data, Sedatives == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Substance_Sedatives_Neg <- calc_n_and_percent(filter(data, Sedatives == "Positive" & ARV.DRM == "Negative"), total_negative)
results$Substance_PolySubstance_Total <- calc_n_and_percent(filter(data, PolySubstance == "Positive"), total_n)
results$Substance_PolySubstance_Pos <- calc_n_and_percent(filter(data, PolySubstance == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Substance_PolySubstance_Neg <- calc_n_and_percent(filter(data, PolySubstance == "Positive" & ARV.DRM == "Negative"), total_negative)

# Stratification of Cannabis Use by PolySubstance
results$Cannabis_with_PolySubstance_Total <- calc_n_and_percent(filter(data, PolySubstance == "Positive" & Cannabis == "Positive"), total_n)
results$Cannabis_with_PolySubstance_Pos <- calc_n_and_percent(filter(data, PolySubstance == "Positive" & Cannabis == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Cannabis_with_PolySubstance_Neg <- calc_n_and_percent(filter(data, PolySubstance == "Positive" & Cannabis == "Positive" & ARV.DRM == "Negative"), total_negative)
results$Cannabis_without_PolySubstance_Total <- calc_n_and_percent(filter(data, PolySubstance == "Negative" & Cannabis == "Positive"), total_n)
results$Cannabis_without_PolySubstance_Pos <- calc_n_and_percent(filter(data, PolySubstance == "Negative" & Cannabis == "Positive" & ARV.DRM == "Positive"), total_positive)
results$Cannabis_without_PolySubstance_Neg <- calc_n_and_percent(filter(data, PolySubstance == "Negative" & Cannabis == "Positive" & ARV.DRM == "Negative"), total_negative)

# Convert results into a data frame
table_df <- data.frame(
  Characteristic = c("Female", "Male", "White", "Asian", "Black","Hispanic", "Multi-Racial", "Native American", "Unknown",
                     "Cannabis", "Cocaine", "Amphetamine", "Opiates", "Sedatives", "Poly-Substance",
                     "Cannabis Use with PolySubstance", "Cannabis Use without PolySubstance"),
  Total_n = sapply(results[c(TRUE, FALSE, FALSE)], function(x) x$n),
  Total_Percent = sapply(results[c(TRUE, FALSE, FALSE)], function(x) x$percent),
  ARV_DRM_Present_n = sapply(results[c(FALSE, TRUE, FALSE)], function(x) x$n),
  ARV_DRM_Present_Percent = sapply(results[c(FALSE, TRUE, FALSE)], function(x) x$percent),
  ARV_DRM_Absent_n = sapply(results[c(FALSE, FALSE, TRUE)], function(x) x$n),
  ARV_DRM_Absent_Percent = sapply(results[c(FALSE, FALSE, TRUE)], function(x) x$percent)
)


 # Calculate n and percent for Cannabis Use stratified by PolySubstance and No PolySubstance
Cannabis_PolySubstance_Total <- calc_n_and_percent(filter(data, Cannabis == "Positive" & PolySubstance == "Positive"), total_n)
Cannabis_PolySubstance_Pos <- calc_n_and_percent(filter(data, Cannabis == "Positive" & PolySubstance == "Positive" & ARV.DRM == "Positive"), total_positive)
Cannabis_PolySubstance_Neg <- calc_n_and_percent(filter(data, Cannabis == "Positive" & PolySubstance == "Positive" & ARV.DRM == "Negative"), total_negative)

Cannabis_NoPolySubstance_Total <- calc_n_and_percent(filter(data, Cannabis == "Positive" & PolySubstance == "Negative"), total_n)
Cannabis_NoPolySubstance_Pos <- calc_n_and_percent(filter(data, Cannabis == "Positive" & PolySubstance == "Negative" & ARV.DRM == "Positive"), total_positive)
Cannabis_NoPolySubstance_Neg <- calc_n_and_percent(filter(data, Cannabis == "Positive" & PolySubstance == "Negative" & ARV.DRM == "Negative"), total_negative)

# Create a separate data frame
cannabis_df <- data.frame(
  Characteristic = c("Cannabis Use with Poly-Substance", "Cannabis Use without Poly-Substance"),
  Total_n = c(Cannabis_PolySubstance_Total$n, Cannabis_NoPolySubstance_Total$n),
  Total_Percent = c(Cannabis_PolySubstance_Total$percent, Cannabis_NoPolySubstance_Total$percent),
  ARV_DRM_Present_n = c(Cannabis_PolySubstance_Pos$n, Cannabis_NoPolySubstance_Pos$n),
  ARV_DRM_Present_Percent = c(Cannabis_PolySubstance_Pos$percent, Cannabis_NoPolySubstance_Pos$percent),
  ARV_DRM_Absent_n = c(Cannabis_PolySubstance_Neg$n, Cannabis_NoPolySubstance_Neg$n),
  ARV_DRM_Absent_Percent = c(Cannabis_PolySubstance_Neg$percent, Cannabis_NoPolySubstance_Neg$percent)
)

table_df <- rbind(table_df,cannabis_df)

#no substance
# Calculate n and percent for No Substance Use
NoSubstance_Total <- calc_n_and_percent(filter(data, Num_of_Substances == 0), total_n)
NoSubstance_Pos <- calc_n_and_percent(filter(data, Num_of_Substances == 0 & ARV.DRM == "Positive"), total_positive)
NoSubstance_Neg <- calc_n_and_percent(filter(data, Num_of_Substances == 0 & ARV.DRM == "Negative"), total_negative)

# Add the No Substance Use row to the separate data frame
nosub_df <-   data.frame(
    Characteristic = "No Substance Use",
    Total_n = NoSubstance_Total$n,
    Total_Percent = NoSubstance_Total$percent,
    ARV_DRM_Present_n = NoSubstance_Pos$n,
    ARV_DRM_Present_Percent = NoSubstance_Pos$percent,
    ARV_DRM_Absent_n = NoSubstance_Neg$n,
    ARV_DRM_Absent_Percent = NoSubstance_Neg$percent
  )

table_df <- rbind(table_df,nosub_df)
#write.csv(table_df,"Table1.csv")
```

## Statistical Analyses

Using Firth logistic regression analyses, we estimated the Odds Ratio (OR) as a measure of the association between substance use and ARV-DRM status. Firth regression was chosen to address the small number of ARV-DRM cases in certain subgroups. 

### First, we tested univariable models for each predictor of interest (substance use) and each other variable considered.

The results of fitting these model are in the p-value column of Table 1

```{r}
data <- data %>%
  mutate(SubstanceUse = ifelse(Num_of_Substances > 0, "TRUE", "FALSE"))
# Create variable of interest vector
VOI <- c("Cannabis","PolySubstance","Amphetamine","Cocaine","Sedatives","Opiates", 
"Race","Age_Enrollment","Sex","Year_of_HIV_Diagnosis","Length_of_Infection","SubstanceUse","IV_Drug_Use")

#Relevel race because Asian is default ref and there is only one participant
data$Race <- relevel(data$Race, ref = "White")
```

```{r}
# Loop through each variable of interest
# Initialize an empty data frame to store results
univar_results_df <- data.frame()

for (var in VOI) {
  sink(tempfile())
  # Fit the model using the updated data
  invisible(firth_model <- logistf(as.formula(paste("ARV.DRM ~", var)), data = data))
  
  # Calculate odds ratios and confidence intervals
  best_model_or <- exp(coef(firth_model))
  conf_int <- exp(confint(firth_model))
  
  # Extract p-values from the model summary
  best_model_summary <- summary(firth_model)
  p_values <- best_model_summary$prob
  
  # Identify the reference level for categorical variables
  ref_level <- if (is.factor(data[[var]])) {
    levels(data[[var]])[1]  # The first level of the factor variable, used as the reference
  } else {
    NA  # For continuous variables, set reference level as NA
  }
  
  # Combine odds ratios, confidence intervals, p-values, and reference level into a data frame
  odds_ratio_df <- data.frame(
    Variable = var,
    Predictor = names(best_model_or),
    Odds_Ratio = round(best_model_or, 3),
    CI_95_Lower = round(conf_int[, 1], 3),
    CI_95_Upper = round(conf_int[, 2], 3),
    P_Value = round(p_values, 5),
    Reference_Level = ref_level  # Include the reference level in the output
  )
  
  # Append the results to the final data frame
  univar_results_df <- rbind(univar_results_df, odds_ratio_df)
  sink()
}

# Filter out rows where Predictor is "(Intercept)"
univar_results_df <- univar_results_df[univar_results_df$Predictor != "(Intercept)", ]
datatable(univar_results_df,rownames = FALSE,caption = "Results of all univariable models")
```

```{r}
#Add univariable analysis of cannabis in data stratified by polysubstance use
poly_data <- data %>% filter(PolySubstance == "Positive") # This compares poly users with cannabis to poly users without
single_data <-  data %>% filter(PolySubstance == "Negative") #This includes cannabis only users to non drug users

firth_model <- logistf(ARV.DRM ~ Cannabis, data = poly_data)  
  # Calculate odds ratios and confidence intervals
  best_model_or <- exp(coef(firth_model))
  conf_int <- exp(confint(firth_model))
  
  # Extract p-values from the model summary
  best_model_summary <- summary(firth_model)
  p_values <- best_model_summary$prob
  
  # Identify the reference level for categorical variables
  ref_level <- if (is.factor(data[[var]])) {
    levels(data[[var]])[1]  # The first level of the factor variable, used as the reference
  } else {
    NA  # For continuous variables, set reference level as NA
  }
  
  # Combine odds ratios, confidence intervals, p-values, and reference level into a data frame
  odds_ratio_df <- data.frame(
    Variable = var,
    Predictor = names(best_model_or),
    Odds_Ratio = round(best_model_or, 3),
    CI_95_Lower = round(conf_int[, 1], 3),
    CI_95_Upper = round(conf_int[, 2], 3),
    P_Value = round(p_values, 5),
    Reference_Level = ref_level  # Include the reference level in the output
  )
  odds_ratio_df$subset <- "polysubstance"
  cannabis_subset_results_df <- odds_ratio_df
  
firth_model <- logistf(ARV.DRM ~ Cannabis, data = single_data)  
  # Calculate odds ratios and confidence intervals
  best_model_or <- exp(coef(firth_model))
  conf_int <- exp(confint(firth_model))
  
  # Extract p-values from the model summary
  best_model_summary <- summary(firth_model)
  p_values <- best_model_summary$prob
  
  # Identify the reference level for categorical variables
  ref_level <- if (is.factor(data[[var]])) {
    levels(data[[var]])[1]  # The first level of the factor variable, used as the reference
  } else {
    NA  # For continuous variables, set reference level as NA
  }
  
  # Combine odds ratios, confidence intervals, p-values, and reference level into a data frame
  odds_ratio_df <- data.frame(
    Variable = var,
    Predictor = names(best_model_or),
    Odds_Ratio = round(best_model_or, 3),
    CI_95_Lower = round(conf_int[, 1], 3),
    CI_95_Upper = round(conf_int[, 2], 3),
    P_Value = round(p_values, 5),
    Reference_Level = ref_level  # Include the reference level in the output
  )
  odds_ratio_df$subset <- "cannabis_only"
  # Append the results to the final data frame
  cannabis_subset_results_df <- rbind(cannabis_subset_results_df, odds_ratio_df)
  cannabis_subset_results_df <- cannabis_subset_results_df[cannabis_subset_results_df$Predictor != "(Intercept)", ]
write.csv(file="Cannabis_univariable_stratified_by_polysubstance.csv",univar_results_df)
```

```{r}
#Add the univariable test results to Table 1 
#Adding the cannabis stratified by polysubstance use was done manually using the result written above
# Standardize names for matching
clean_name <- function(x) {
  x %>%
    tolower() %>%
    gsub("[[:space:]-]", "", .) %>% # Remove spaces, hyphens
    gsub("positive", "", .) %>%     # Remove "Positive" suffix
    gsub("use", "", .)  %>%            # Remove "Use" if applicable
    gsub("race", "", .) %>%
    gsub("sex", "", .) %>%
    gsub("_", "", .) 
    }

# Clean both Characteristic and Predictor columns
table_df <- table_df %>%
  mutate(Cleaned_Characteristic = clean_name(Characteristic))

univar_results_df <- univar_results_df %>%
  mutate(Cleaned_Predictor = clean_name(Predictor))

# Add the p-value column
table_df <- table_df %>%
  left_join(
    univar_results_df %>% select(Cleaned_Predictor, P_Value),
    by = c("Cleaned_Characteristic" = "Cleaned_Predictor")
  ) %>%
  select(-Cleaned_Characteristic) # Remove intermediate column if unnecessary

# Add an asterisk for missing data
variables_with_na <- names(data)[sapply(data, function(x) any(is.na(x)))]

table_df <- table_df %>%
  mutate(
    Characteristic = ifelse(
      gsub("[[:space:]-]", "", Characteristic) %in% clean_name(variables_with_na),
      paste0(Characteristic, "*"),
      Characteristic
    )
  )


# 2. Add an asterisk to any value of Characteristic if it has missing data in `data`
# Create a list of variables in data with missing values
variables_with_na <- names(data)[sapply(data, function(x) any(is.na(x)))]

# Add an asterisk to the Characteristic column in table_df if it matches any of the variables with missing data
table_df <- table_df %>%
  mutate(
    Characteristic = ifelse(
      gsub("_.*", "", Characteristic) %in% variables_with_na, 
      paste0(Characteristic, "*"), 
      Characteristic
    )
  )
```

### Second, we tested models including combinations of those predictors that were significant at the p < 0.20 level. 

```{r}
library(logistf)

# Variables to consider for model combinations
LowPVars <- c("Cannabis", "Length_of_Infection", "Sex", "Race","IV_Drug_Use")

# Function to fit the model using Firth logistic regression and store results
fit_and_store_model <- function(formula_str, data_factor) {
  model <- logistf(as.formula(formula_str), data = data_factor)
  
  # Extract coefficients and p-values
  coef_names <- names(model$coefficients)
  coefficients <- model$coefficients
  p_values <- model$prob
  
  # Remove the intercept from significant factors
  significant_factors <- coef_names[p_values < 0.05 & coef_names != "(Intercept)"]
  significant_pvals <- p_values[p_values < 0.05 & coef_names != "(Intercept)"]
  
  # Calculate the likelihood ratio test statistic and p-value manually
  loglik_full <- model$loglik["full"]
  loglik_null <- model$loglik["null"]
  lrt_statistic <- -2 * (loglik_null - loglik_full)
  lrt_p_value <- pchisq(lrt_statistic, df = 1, lower.tail = FALSE)
  
  return(list(formula = formula_str, LRT_P_Value = lrt_p_value, 
              significant_factors = significant_factors, p_values = significant_pvals, model = model))
}

# Create a list to store the models and their results
model_comparisons <- list()

# Generate all possible combinations of LowPVars
for (k in 1:length(LowPVars)) {
  combinations <- combn(LowPVars, k, simplify = FALSE)
  
  for (combo in combinations) {
    # Create the formula string
    formula_str <- paste("ARV.DRM ~", paste(combo, collapse = " + "))
    
    # Fit the model and store the results
    #model_name <- paste("Model with", paste(combo, collapse = ", "))
    model_name <- formula_str
    model_comparisons[[model_name]] <- fit_and_store_model(formula_str, data)
  }
}

# Create a data frame to store results for all models
model_results <- data.frame(
  Model = names(model_comparisons),
  LRT_P_Value = sapply(model_comparisons, function(x) round(x$LRT_P_Value, 5)),
  Significant_Factors = sapply(model_comparisons, function(x) ifelse(length(x$significant_factors) == 0, 
                                                                     "None", 
                                                                     paste(x$significant_factors, collapse = ", "))),
  P_Values = sapply(model_comparisons, function(x) ifelse(length(x$p_values) == 0, 
                                                          "None", 
                                                          paste(round(x$p_values, 5), collapse = ", ")))
)


# Use datatable() to create an interactive table
datatable(model_results, options = list(pageLength = 10), rownames = FALSE)
write.csv(file="Table_S2.csv",model_results)
```

### Results of the best model including cannabis

```{r}
# Fit the model using Firth logistic regression
model <- logistf(ARV.DRM ~ Cannabis + Sex + Race, data = data)

# Use summary() to get detailed results
model_summary <- summary(model)

tbl <- logistf_tablize(model_summary)
datatable(tbl,rownames = FALSE)
cannabis_in_best_multivariable <- tbl %>% filter(Variable=="CannabisPositive")
cannabis_in_best_multivariable$Model <- "ARV.DRM ~ Cannabis + Sex + Race"
cannabis_in_best_multivariable$Sample <- "Full sample"
```

### Results of the best model including IV drug use

```{r}
# Fit the model using Firth logistic regression
model <- logistf(ARV.DRM ~ Length_of_Infection + Race + IV_Drug_Use,data=data)

# Use summary() to get detailed results
model_summary <- summary(model)

tbl <- logistf_tablize(model_summary)
datatable(tbl,rownames = FALSE)
IV_in_best_multivariable <- tbl %>% filter(Variable=="IV_Drug_Use")
IV_in_best_multivariable$Model <- "ARV.DRM ~ Length_of_Infection + Race + IV_Drug_Use"
IV_in_best_multivariable$Sample <- "IV drug use responders"
```

### Consider the interaction between cannabis and polysubstance use

Because we observed a trend in our data that participants who test positive for cannabis tend to test positive for other substances, we tested this relationship using Pearson's Chi-squared test and found a highly significant correlation. 

```{r}
# Create a contingency table
table_data <- table(data$Cannabis, data$PolySubstance)
# Perform a Chi-Square test
chi_square_test <- chisq.test(table_data)
print(chi_square_test)
library(psych)
# Calculate the phi correlation
phi_correlation <- phi(table_data)
print(phi_correlation)

df_table <- as.data.frame(table_data)

# Rename columns for better readability
colnames(df_table) <- c("Cannabis", "PolySubstance", "Frequency")

# Create a dot plot
ggplot(df_table, aes(x = Cannabis, y = PolySubstance)) +
  geom_point(aes(size = Frequency), color = "blue", alpha = 0.7) +
  scale_size_continuous(range = c(3, 13)) +  # Adjust the range of dot sizes
  labs(
    x = "Cannabis Use",
    y = "PolySubstance Use",
    size = "Number of Observations",
    title = "Dot Plot of Cannabis and PolySubstance Use"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(color = "gray80", linetype = "dashed")
  )
```

## Consider the interaction between IV drug use and polysubstance use

We tested this relationship using Pearson's Chi-squared test and did not find a significant correlation

```{r}
# Create a contingency table
table_data <- table(data$IV_Drug_Use, data$PolySubstance)
# Perform a Chi-Square test
chi_square_test <- chisq.test(table_data)
print(chi_square_test)
library(psych)
# Calculate the phi correlation
phi_correlation <- phi(table_data)
print(phi_correlation)

df_table <- as.data.frame(table_data)

# Rename columns for better readability
colnames(df_table) <- c("IV_Drug_Use", "PolySubstance", "Frequency")

# Create a dot plot
ggplot(df_table, aes(x = IV_Drug_Use, y = PolySubstance)) +
  geom_point(aes(size = Frequency), color = "blue", alpha = 0.7) +
  scale_size_continuous(range = c(3, 13)) +  # Adjust the range of dot sizes
  labs(
    x = "IV Drug Use",
    y = "PolySubstance Use",
    size = "Number of Observations",
    title = "Dot Plot of IV Drug Use and PolySubstance Use"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(color = "gray80", linetype = "dashed")
  )
```

## Cannabis and polysubstance use

To determine whether PolySubstance use is a confounder in the relationship between Cannabis use and ARV.DRM, we evaluated whether the inclusion of PolySubstance use in a regression model significantly changes the association between Cannabis use and ARV.DRM. A confounder is a variable that is associated with both the exposure (Cannabis use) and the outcome (ARV.DRM) and could distort the observed relationship.

```{r}
# Model with both Cannabis and PolySubstance use, controlling for sex and race
cannabis_multivar_with_poly <- logistf(ARV.DRM ~ Cannabis + PolySubstance + Sex + Race, data = data)

# Compare coefficients and p-values for Cannabis use in both models
datatable(logistf_tablize(summary(cannabis_multivar_with_poly)),rownames = FALSE,caption="Results of model with polysubstance use")
cannabis_in_best_with_poly <- logistf_tablize(summary(cannabis_multivar_with_poly))%>%filter(Variable=="CannabisPositive")
cannabis_in_best_with_poly$Model <- "ARV.DRM ~ Cannabis + PolySubstance + Sex + Race"
cannabis_in_best_with_poly$Sample <- "Full sample"
```

## IV Drug Use and polysubstance use

Check it in case

```{r}
# Model with both IV_Drug_Use and PolySubstance use, controlling for sex and race
IV_Drug_Use_multivar_with_poly <- logistf(ARV.DRM ~ IV_Drug_Use + PolySubstance + Length_of_Infection + Race, data = data)

# Compare coefficients and p-values for IV_Drug_Use use in both models
datatable(logistf_tablize(summary(IV_Drug_Use_multivar_with_poly)),rownames = FALSE,caption="Results of model with polysubstance use")
IV_Drug_Use_in_best_with_poly <- logistf_tablize(summary(IV_Drug_Use_multivar_with_poly))%>%filter(Variable=="IV_Drug_Use")
IV_Drug_Use_in_best_with_poly$Model <- "ARV.DRM ~ IV_Drug_Use + PolySubstance + Length_of_Infection + Race"
IV_Drug_Use_in_best_with_poly$Sample <- "IV Drug Use Responses"
```

### Interaction between polysubstance use and cannabis use as a supressor variable

Because including polysubstance in the model decreases the p-value and the odds-ratio for cannabis, we further considered whether there is an interaction between cannabis and polysubstance use, i.e., does the effect of being a cannabis user on ARV.DRM change depending on whether or not the participant is also a polysubstance user.

We find evidence for polysubstance use and its interaction with cannabis as a suppressor variable: A variable that does not have a strong relationship with the outcome variable, but is related to another predictor, in this case, cannabis

```{r}
# Model with interaction term between Cannabis and PolySubstance, controlling for sex and race
interaction_model <- logistf(ARV.DRM ~ Cannabis * PolySubstance + Sex + Race, data = data)
datatable(logistf_tablize(summary(interaction_model)),rownames=FALSE,caption="Results of model with interaction between cannabis and polysubstance use")
cannabis_in_best_with_polyinteraction <- logistf_tablize(summary(interaction_model))%>%filter(Variable=="CannabisPositive")
cannabis_in_best_with_polyinteraction$Model <- "ARV.DRM ~ Cannabis * PolySubstance + Sex + Race"
cannabis_in_best_with_polyinteraction$Sample <- "Full sample"
```

### Stratify the participants by polysubstance use status

To understand how polysubstance use affects the relationship between cannabis use and ARV.DRM, we ran the best model on each level of polysubstance use.

Our finding that cannabis use only has a significant effect on ARV.DRM within non-polysubstance users suggests that the documented negative effect of other drugs of abuse on HIV outcomes (not observed in this study) as well as the association between non-cannabis substance use and other higher-risk behaviors and characteristics (e.g. needle sharing, condomless sex, homelessness) may prevents the positive association between cannabis use and absence of ARV.DRM that is seen in non-polysubstance users.

```{r}
poly_data <- data %>% filter(PolySubstance == "Positive")
poly_model <- logistf(ARV.DRM ~ Cannabis + Sex + Race, data = poly_data)
datatable(logistf_tablize(summary(poly_model)),rownames=FALSE,caption="Results of best model with only polysubstance users")
single_data <-  data %>% filter(PolySubstance == "Negative")
single_model <- logistf(ARV.DRM ~ Cannabis + Sex + Race, data = single_data)
datatable(logistf_tablize(summary(single_model)),rownames=FALSE,caption="Results of best model with only non polysubstance users")
cannabis_in_poly_subset <- logistf_tablize(summary(poly_model))%>%filter(Variable=="CannabisPositive")
cannabis_in_poly_subset$Model <- "ARV.DRM ~ Cannabis"
cannabis_in_poly_subset$Sample <- "Polysubstance use participants"
cannabis_in_nonpoly_subset <- logistf_tablize(summary(single_model))%>%filter(Variable=="CannabisPositive")
cannabis_in_nonpoly_subset$Model <- "ARV.DRM ~ Cannabis"
cannabis_in_nonpoly_subset$Sample <- "Cannabis only use participants"
```

```{r}
#Combine all the cannabis results
cannabis_dataframes <- ls(pattern = "^cannabis_in")

# Combine all identified data frames using rbind
cannabis_results <- do.call(rbind, mget(cannabis_dataframes))

cannabis_results <- cannabis_results %>%
  mutate(
    `95% CI (OR)` = paste0(round(CI_Lower, 2), ", ", round(CI_Upper, 2))
  ) %>%
  select(-CI_Lower, -CI_Upper) # Remove the original columns
write.csv(file="Table2.csv",cannabis_results)
```

It is helpful to visualize the data categories to understand the limitations of this analysis, mainly the relatively small number of ARV-DRM positive study participants limits the size of the categories.

```{r,fig.dim=c(8.8)}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Create the data frame using dplyr
df <- data %>%
  group_by(Cannabis, PolySubstance, ARV.DRM) %>%
  summarize(Frequency = n(), .groups = "drop")

# Ensure the factors are in the correct order for plotting
df$Cannabis <- factor(df$Cannabis, levels = c("Negative", "Positive"))
df$PolySubstance <- factor(df$PolySubstance, levels = c("Negative", "Positive"))
# Ensure ARV.DRM is a factor with levels in the correct order
df$ARV.DRM <- factor(df$ARV.DRM, levels = c("Positive", "Negative"))

# Create the bubble plot with adjusted spacing
bubble_plot <- ggplot(df, aes(x = Cannabis, y = PolySubstance)) +
  geom_point(aes(
    size = Frequency,
    fill = ifelse(PolySubstance == "Negative" & ARV.DRM == "Negative", "red", as.character(ARV.DRM))
  ), shape = 21, color = "black", alpha = 0.7)+
  geom_text(aes(
    label = paste0("n=", Frequency),
    hjust = ifelse(ARV.DRM == "Negative",
                   ifelse(Cannabis == "Negative", 2.8, -2),  # Move left for Cannabis Negative, right for Cannabis Positive
                   0.5),  # Center for ARV.DRM Positive
    vjust = ifelse(Frequency == 1, -0.5, ifelse(ARV.DRM == "Positive", -2, 0.5))  # Move up for all ARV.DRM Positive, center for ARV.DRM Negative
  ), size = 3) +
  scale_size_continuous(range = c(1, 50),breaks = c(1, 10, 150, 200), guide = guide_legend(
    title = "Number of Participants")) +
  scale_fill_manual(
    values = c("Positive" = "black", "Negative" = "lightgray", "red" = "red"),
    labels = c("Positive", "Negative", "Negative (Polysubstance)"),
    guide = guide_legend(title = "ARV.DRM"))+
  scale_x_discrete(expand = expansion(mult = c(1.5,1.5))) +  # Adjust horizontal spacing
  scale_y_discrete(expand = expansion(mult = c(1.5,1.5))) +  # Adjust vertical spacing
  labs(
    x = "Cannabis Use",
    y = "Polysubstance Use",
    title = "Participants' Cannabis Use, Polysubstance Use, and ARV-DRM status",
    subtitle = "Bubble size indicates the number of observations"
  ) +
  annotate("text", x = 1.5, y = -0, label = "ARV-DRM ~ Cannabis + Sex + Race (in non polysubstance users)\nCannabis p-value: 0.025\nCannabis odds ratio (relative to no cannabis use): 0.19",
           color = "red", size = 4, hjust = 0.5) +
  theme_minimal() +
theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),  # Increase legend title size
    legend.text = element_text(size = 10),  # Increase legend text size
    axis.text.x = element_text(size = 14),  # Increase x-axis label size
    axis.text.y = element_text(size = 14),  # Increase y-axis label size
    axis.title.x = element_text(size = 16, margin = margin(t = 10)),  # Increase x-axis title size
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),  # Increase y-axis title size
    plot.title = element_text(size = 18, hjust = 0.5),  # Increase title size
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    panel.background = element_rect(fill = "white", color = NA),  # Set panel background to white
    plot.background = element_rect(fill = "white", color = NA)  # Increase subtitle size
  )
bubble_plot
ggsave("bubble_plot.png", bubble_plot, width = 10, height = 7, dpi = 300)
```

Because firth binomial regression already accounts for the very small sample size in one of our cases, we provide an additional test for additional support and again find a significant result.

```{r}
# Create a contingency table
table_data <- table(data$Cannabis, data$ARV.DRM, data$PolySubstance)

# Perform Fisher's Exact Test on a specific subset, e.g., only the cases without polysubstance use
# Subset the data for cases with PolySubstance == "Negative"
subset_data <- data %>% filter(PolySubstance == "Negative")

# Create a contingency table for the subset
contingency_table <- table(subset_data$Cannabis, subset_data$ARV.DRM)

# Run Fisher's Exact Test
fisher_test_result <- fisher.test(contingency_table)
print(fisher_test_result)
```

We perform a sensitivity to test to test how sensitive our analysis is to adding additional cases of ARV-DRM positive cannabis users to the data subset of non polysubstance users and find that if even one additional ARV-DRM positive and cannabis positive user was included in the dataset, the p-value of cannabis is only marginally significant.

```{r}
# Load necessary packages
library(dplyr)
library(logistf)

# Create proportional additions based on the existing dataset
# Calculate proportions for each factor
cannabis_props <- table(single_data$Cannabis) / nrow(single_data)
polysubstance_props <- table(single_data$PolySubstance) / nrow(single_data)
sex_props <- table(single_data$Sex) / nrow(single_data)
race_props <- table(single_data$Race) / nrow(single_data)

# Specify the number of new participants to add
n_new <- 43  # Adjust this as needed

# Simulate new participants proportionally
set.seed(42)  # For reproducibility
new_rows <- data.frame(
  Cannabis = sample(names(cannabis_props), size = n_new, replace = TRUE, prob = cannabis_props),
  PolySubstance = sample(names(polysubstance_props), size = n_new, replace = TRUE, prob = polysubstance_props),
  ARV.DRM = sample(c("Positive", "Negative"), size = n_new, replace = TRUE, prob = c(0.05, 0.95)), # Sample proportionally
  Sex = sample(names(sex_props), size = n_new, replace = TRUE, prob = sex_props),
  Race = sample(names(race_props), size = n_new, replace = TRUE, prob = race_props)
)

# Ensure single_data has only the columns that are in new_rows
common_cols <- intersect(names(data), names(new_rows))
data_subset <- data[, common_cols]

# Combine the subsetted single_data with new_rows
updated_data <- bind_rows(data_subset, new_rows)
updated_data$ARV.DRM <- as.factor(updated_data$ARV.DRM)

# Proceed with the analysis
firth_model_updated <- logistf(ARV.DRM ~ Cannabis*PolySubstance + Sex + Race, data = updated_data)

# Print the summary of the updated model
model_summary <- summary(firth_model_updated)

# Create a data frame for reporting
report_df <- data.frame()
  Variable = names(model_summary$coefficients)[-1]  # Exclude the intercept
  P_Value = model_summary$prob[-1]

# Create a data frame for reporting
report_df <- data.frame(
  Variable = names(model_summary$coefficients)[-1],  # Exclude the intercept
  P_Value = model_summary$prob[-1],  # P-values
  Odds_Ratio = exp(model_summary$coefficients[-1]),  # Odds Ratios
  CI_Lower = exp(model_summary$ci.lower[-1]),  # Lower bound of 95% CI
  CI_Upper = exp(model_summary$ci.upper[-1])   # Upper bound of 95% CI
)

# Round values for clarity
report_df <- report_df %>%
  mutate(
    P_Value = round(P_Value, 4),
    Odds_Ratio = round(Odds_Ratio, 2),
    CI_Lower = round(CI_Lower, 2),
    CI_Upper = round(CI_Upper, 2)
  )

# Display the updated results
print(report_df)

```

## CSF

```{r}
CSF <- read.csv("data/TableS3.csv")

library(janitor)

# Sanitize column names
CSF <- CSF %>%
  janitor::clean_names()

# Convert the percentage values to numeric
CSF <- CSF %>%
  mutate(csf_escape_percent_of_visits = as.numeric(gsub("%", "", csf_escape_csf_vl_plasma_vl_percent_of_visits)))

# Create a numeric encoding for Urinalysis (0 = None, 1 and 2 = Substance use)
CSF <- CSF %>%
  mutate(urinalysis_numeric = case_when(
    urinalysis == "None" ~ 0,
    TRUE ~ 1 # Combine all substance use into 1
  ))

t_test <- t.test(
  csf_escape_percent_of_visits ~ urinalysis_numeric,
  data = CSF,
  var.equal = TRUE # Use var.equal = FALSE if variances differ significantly
)

# Print results
cat("T-test results:\n")
cat("Mean CSF Escape (% of visits) for Non-Users:", 
    mean(CSF$csf_escape_percent_of_visits[CSF$urinalysis_numeric == 0], na.rm = TRUE), "\n")
cat("Mean CSF Escape (% of visits) for Users:", 
    mean(CSF$csf_escape_percent_of_visits[CSF$urinalysis_numeric == 1], na.rm = TRUE), "\n")
cat("P-value for the t-test:", round(t_test$p.value, 4), "\n")
```

```{r}
# Recode polysubstance use based on cannabis use
data$poly_by_cannabis <- with(data, ifelse(
  PolySubstance == "Negative" & Num_of_Substances == 0, 
  "No drug use", 
  ifelse(PolySubstance == "Negative" & Num_of_Substances == 1, 
         "Single drug use", 
         ifelse(Cannabis == "Positive", 
                "Positive_with_cannabis", 
                "Positive_without_cannabis")
  )
))

# Set "No drug use" as the reference level
data$poly_by_cannabis <- factor(data$poly_by_cannabis, levels = c("No drug use", "Single drug use", "Positive_with_cannabis", "Positive_without_cannabis"))

# Run the logistf model
logistf_model <- logistf(ARV.DRM ~ poly_by_cannabis, data = data)

logistf_results <- logistf_tablize(summary(logistf_model))

# View the results
print(logistf_results)
```

```{r}
library(curl)
# Path to save the downloaded file
output_file <- "CSF_escape.xlsx"
curl_download("https://uflorida-my.sharepoint.com/:x:/g/personal/pinkp0wr_ufl_edu/EZO0YD_hed1Btd-XGYab2_QBkxja10mZmnay3wMRkp7PiQ?e=9uYdm3",output_file)
CSF <- read_excel("CSF_escape.xlsx",sheet=3)
getwd()
```

CSF Escape 

```{r}
library(readxl)
# Define the file path
file_path <- "original_data/Substance Use & CSF Escape.xlsx"

# Get the names of all sheets in the Excel file
sheet_names <- excel_sheets(file_path)

# Read each sheet into a data frame and store in a list, naming each element by its sheet name
data_frames <- lapply(sheet_names, function(sheet) {
  read_excel(file_path, sheet = sheet)
})
names(data_frames) <- sheet_names

library(dplyr)

# Extract the relevant data frame
df <- data_frames[["Urinalysis & VL Combined"]]

# Fix the typo in the column name
colnames(df)[colnames(df) == "CSF Ecsape with Plasma Suppression"] <- "CSF Escape with Plasma Suppression"

# Summarize the data
summary_df <- df %>%
  group_by(PATID) %>%
  summarize(
    `CSF Escape` = ifelse(any(`CSF Escape` == 1), 1, 0),
    `CSF Escape with Plasma Suppression` = ifelse(any(`CSF Escape with Plasma Suppression` == 1), 1, 0)
  )

orig_data <- read.csv("original_data/Hale2023_originalData.csv")

summary_df <- merge(summary_df,orig_data,by="PATID")

library(logistf)
# Define models for Set 1
model_1 <- logistf(`CSF Escape` ~ ARVDRM, data = summary_df)
model_2 <- logistf(`CSF Escape` ~ Cannabis, data = summary_df)
model_3 <- logistf(`CSF Escape` ~ Cannabis + Poly_Substance, data = summary_df)
model_4 <- logistf(`CSF Escape` ~ Cannabis * Poly_Substance, data = summary_df)
model_5 <- logistf(`CSF Escape with Plasma Suppression` ~ Cannabis, data = summary_df)
model_6 <- logistf(`CSF Escape with Plasma Suppression` ~ Cannabis + Poly_Substance, data = summary_df)
model_7 <- logistf(`CSF Escape with Plasma Suppression` ~ Cannabis * Poly_Substance, data = summary_df)

# Define models for Set 2
model_8 <- logistf(ARVDRM ~`CSF Escape`, data = summary_df)
model_9 <- logistf(ARVDRM ~`CSF Escape` + Cannabis + Poly_Substance, data = summary_df)
model_10 <- logistf(ARVDRM ~`CSF Escape` + Cannabis * Poly_Substance, data = summary_df)
model_11 <- logistf(ARVDRM ~ `CSF Escape with Plasma Suppression` + Cannabis, data = summary_df)
model_12 <- logistf(ARVDRM ~ `CSF Escape with Plasma Suppression` + Cannabis + Poly_Substance, data = summary_df)
model_13 <- logistf(ARVDRM ~ `CSF Escape with Plasma Suppression` + Cannabis * Poly_Substance, data = summary_df)

# Store models in a list for easy access
models <- list(
  model_1, model_2, model_3, model_4, model_5, model_6,
  model_7, model_8, model_9, model_10, model_11, model_12,model_13
)
summary <- summary(model_8)
logistf_tablize(summary)

# Define a function to extract and process each model
process_model <- function(model, model_index) {
  # Extract model summary
  model_summary <- summary(model)
  
  # Tabulate the results using your custom function
  tabulated <- logistf_tablize(model_summary)
  
  # Add columns for model name and outcome
  tabulated$Model <- paste0("Model ", model_index, ": ", deparse(formula(model)))
  tabulated$Outcome <- as.character(formula(model)[[2]])  # Extract left-hand side of the formula
  
  return(tabulated)
}

# Apply the function to all models and combine results
model_results <- do.call(rbind, lapply(seq_along(models), function(i) {
  process_model(models[[i]], i)
}))

# View the final combined data frame
head(model_results)

#Amphetamine (Add this to the text)
Amphetamine_CSF <- summary(logistf(`CSF Escape` ~ Amphetamine, data = summary_df))
logistf_tablize(Amphetamine_CSF)
Amphetamine_ARVDRM <- summary(logistf(ARVDRM ~ Amphetamine, data = summary_df))
logistf_tablize(Amphetamine_ARVDRM)
```

High viremia

```{r}
# Cannabis and viral load
# Make a high viremia variable per vist
df <- df %>%
  mutate(high_viremia = ifelse(`Plasma Viral Load` > 1000, 1, 0))

#Add poly
df <- df %>% mutate(PolySubstance = ifelse(`# of Substances`>1,1,0))
library(lme4)

model_with_Poly <- glmer(high_viremia ~ Marijuana*PolySubstance + (1 | PATID), 
               data = df, family = binomial)
summary(model_with_Poly)
# Extract the fixed-effect coefficient
exp(fixef(model_with_Poly)["Marijuana"])

model_without_Poly <- glmer(high_viremia ~ Marijuana + (1 | PATID), 
               data = df, family = binomial)
summary(model_without_Poly)
```

```{r}
#Subset to remove all polysubstance users
df_marijuana_or_none <- df %>%
  filter(Marijuana %in% c(0, 1) &  # Include only rows with Marijuana = 0 or 1
           Cocaine == 0 &          # Ensure no use of other substances
           Opiates == 0 &
           Meth == 0 &
           Amphetamines == 0 &
           Sedatives == 0 &
           PCP == 0)

model_marijuana_or_none <- glmer(high_viremia ~ Marijuana + (1 | PATID), 
     data = df_marijuana_or_none, family = binomial)
summary(model_marijuana_or_none)

library(ggplot2)
# Summarize data
summary_df <- df_marijuana_or_none %>%
  group_by(Marijuana) %>%
  summarize(Proportion_High_Viremia = mean(high_viremia), Count = n())
ggplot(df_marijuana_or_none, aes(x = as.factor(high_viremia), fill = as.factor(Marijuana))) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportion of Marijuana Use in High vs Low Viremia",
    x = "High Viremia (0 = No, 1 = Yes)",
    y = "Proportion",
    fill = "Marijuana Use"
  )
```

```{r}
# Calculate odds ratio and 95% confidence intervals for glmer model
coefs <- summary(model_marijuana_or_none)$coefficients
z_score <- 1.96  # For 95% confidence interval

# Calculate odds ratio and confidence intervals
OR <- exp(coefs[, "Estimate"])
CI_lower <- exp(coefs[, "Estimate"] - z_score * coefs[, "Std. Error"])
CI_upper <- exp(coefs[, "Estimate"] + z_score * coefs[, "Std. Error"])

# Format results in a data frame
OR_results <- data.frame(
  Variable = rownames(coefs),
  Odds_Ratio = round(OR, 2),
  CI_Lower = round(CI_lower, 2),
  CI_Upper = round(CI_upper, 2)
)

# Print results
print(OR_results)
```