---
title: "Analysis of interaction between specific substance use and poly substance use"
author: "Heather Kates"
date: "2024-09-20"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r,echo=FALSE}
#Load required libraries
library(dplyr)
library(logistf)
library(ggplot2)
library(tidyr)
select <- dplyr::select
```

### The original dataset is used with no subsetting or re-coding

This data is cols AU:BB from sheet "Master Sheet" in https://www.dropbox.com/scl/fi/j37pkkg8624byolqf3aim/MPH-Dataset-Stata.xlsx?rlkey=hdsylsmkursvzazd7a8yc9tk7&st=n00hxd0w&dl=0

```{r}
# Read in the data
data <- read.csv("original_data/Hale2023_originalData.csv")
colnames(data) <- gsub("\\.","_",colnames(data))
colnames(data) <- gsub("__","_",colnames(data))
data <- na.omit(data)
# Convert binary variables (substance use) to factors
data_factor <- data %>%
  mutate(across(c(ARVDRM, Cannabis, Amphetamine, Cocaine, Sedatives, Opiates, Phencyclidine, Poly_Substance),
                ~ factor(.x, levels = c(0, 1), labels = c(0, 1))))
#write.csv(data_factor,"data/TableS1.csv",row.names = FALSE)
```

Read in demographic data and merge with data
This data is cols B:O from sheet "Master Sheet" in https://www.dropbox.com/scl/fi/j37pkkg8624byolqf3aim/MPH-Dataset-Stata.xlsx?rlkey=hdsylsmkursvzazd7a8yc9tk7&st=n00hxd0w&dl=0

```{r}
master_data <- read.csv("original_data/Hale2023_MasterData.csv")
master_data$Cannabis <- as.factor(master_data$Cannabis)
#omit participants without urinalysis results
master_data <- master_data %>% filter(!is.na(Cannabis))
master_data <- master_data %>% select_if(~!all(is.na(.)))
```

# Create a single race variable

```{r}
# Create a new "Race" variable based on the binary race columns
master_data$Race <- ifelse(master_data$MultiRacial == 1, "MultiRacial",
                                ifelse(master_data$Asian == 1, "Asian",
                                ifelse(master_data$Black == 1, "Black",
                                ifelse(master_data$Native_American == 1, "Native_American",
                                ifelse(master_data$Native_Hawaiin == 1, "Native_Hawaiin",
                                ifelse(master_data$White == 1, "White", "Unknown"))))))

# Convert "Race" to a factor
master_data$Race <- as.factor(master_data$Race)
```

# Summary of the data

```{r}
# Load necessary library
library(dplyr)

# Function to summarize only binary variables in the data frame
summarize_binary_variables <- function(data) {
  # Remove columns that are entirely NA
  data <- data %>% select_if(~!all(is.na(.)))

  # Initialize an empty data frame for the summary
  summary_data <- data.frame(Variable = character(),
                             Level = character(),
                             n = numeric(),
                             Percentage = numeric(),
                             stringsAsFactors = FALSE)

  # Loop through each column to check if it's binary
  for (var in names(data)) {
    unique_vals <- unique(na.omit(data[[var]]))
    
    # Check if the variable is binary (only two unique values, e.g., 0 and 1)
    if (length(unique_vals) == 2) {
      for (level in unique_vals) {
        count <- sum(data[[var]] == level, na.rm = TRUE)
        total <- sum(!is.na(data[[var]]))
        percentage <- round((count / total) * 100, 2)
        summary_data <- rbind(summary_data, data.frame(Variable = var, Level = as.character(level), n = count, Percentage = percentage))
      }
    }
  }

  return(summary_data)
}

# Use the function on your data frame
summary_result <- summarize_binary_variables(master_data)

# Print the summary result
print(summary_result)
master_data_restore <- master_data

```

# Visualize data faceted by ARV-DRM

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(patchwork)

# Rename and recode variables as needed
master_data <- master_data %>%
  dplyr::rename(Sex = Gender) %>%
  mutate(Race = dplyr::recode(Race,Native_American="Native\nAmerican")) %>%
  mutate(Race = dplyr::recode(Race,MultiRacial="Multi-\nracial")) %>%
  mutate(
    Sex = dplyr::recode(Sex, `2` = "Male", `1` = "Female"),
    Cannabis = factor(Cannabis, levels = c(0, 1), labels = c("Negative", "Positive")),
    Amphetamine = factor(Amphetamine, levels = c(0, 1), labels = c("Negative", "Positive")),
    Cocaine = factor(Cocaine, levels = c(0, 1), labels = c("Negative", "Positive")),
    Sedatives = factor(Sedatives, levels = c(0, 1), labels = c("Negative", "Positive")),
    Opiates = factor(Opiates, levels = c(0, 1), labels = c("Negative", "Positive")),
    Phencyclidine = factor(Phencyclidine, levels = c(0, 1), labels = c("Negative", "Positive")),
    PolySubstance = factor(PolySubstance, levels = c(0, 1), labels = c("Negative", "Positive")),
    CSF_Escape = factor(CSF_Escape, levels = c(0, 1), labels = c("Negative", "Positive")),
    ARV.DRM = factor(ARV.DRM, levels = c(0, 1), labels = c("Absent", "Present")),
    Sex = factor(Sex, levels = c("Female", "Male"))
  )

# Load tidyr for complete() function
library(tidyr)
# Function to create donut charts with n values embedded in labels and slice labels
create_donut_chart_with_labels <- function(var) {
  # Calculate n values and update levels to include n
  df <- master_data %>%
    group_by(.data[[var]]) %>%
    mutate(
      !!var := paste0(.data[[var]], " (n = ", n(), ")")
    ) %>%
    ungroup() %>%
    group_by(.data[[var]], ARV.DRM) %>%
    summarise(Count = n(), .groups = 'drop') %>%
    group_by(.data[[var]]) %>%
    mutate(Percentage = Count / sum(Count) * 100,
           Label = paste0("n = ", Count))  # Create labels for the slices
  
  # Determine the title text based on the variable
  title_text <- if (var %in% c("Sex", "CSF_Escape")) {
    paste("ARV.DRM status within", var)
  } else {
    paste("ARV.DRM status within", var, "use")
  }
  
  # Generate the donut chart
  ggplot(df, aes(x = 2, y = Percentage, fill = ARV.DRM)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    xlim(0.5, 2.5) +  # Creates the hollow center for the donut chart
    scale_fill_manual(values = c("Absent" = "blue", "Present" = "orange")) +
    facet_wrap(~ .data[[var]]) +
    # Add slice labels using geom_text
    # Use geom_label to add a white box around the text
    geom_label(aes(x=1.5,label = Label), position = position_stack(vjust = 0.8), size = 2, fill = "white",hjust=0) +
    labs(
      title = title_text,
      x = NULL, y = NULL
    ) +
    theme_void() +
    theme(
      plot.title = element_text(size = 10),
      legend.position = "right"
    )
}

# List of binary variables
binary_variables <- c("Sex", "Cannabis", "Amphetamine", "Cocaine", "Sedatives", "Opiates", "Phencyclidine", "PolySubstance", "CSF_Escape")
# Generate donut charts for all binary variables
donut_charts_by_arv <- lapply(binary_variables, create_donut_chart_with_labels)

# Combine all pie charts in a grid
pie_charts_grid <- wrap_plots(donut_charts_by_arv, ncol = 3)

# Create histograms for continuous variables
hist_age <- ggplot(master_data, aes(x = Age_Enrollment, fill = ARV.DRM)) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.7) +
  facet_wrap(~ ARV.DRM) +
  scale_fill_manual(values = c("Absent" = "blue", "Present" = "orange"))+
  labs(title = "Age at Enrollment by ARV.DRM Status", x = "Age", y = "Count") +
  theme_minimal()+
    theme(plot.title = element_text(size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),axis.title.x = element_text(size = 8),
    plot.margin = margin(10, 5, 10, 10)) # Decrease right margin)

hist_hiv_diagnosis <- ggplot(master_data, aes(x = Year_of_HIV_Diagnosis, fill = ARV.DRM)) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.7) +
  facet_wrap(~ ARV.DRM) +
  scale_fill_manual(values = c("Absent" = "blue", "Present" = "orange"))+
  labs(title = "Year of HIV Diagnosis by ARV.DRM Status (n=207)", x = "Year of HIV Diagnosis", y = "Count") +
  theme_minimal()+
    theme(plot.title = element_text(size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title.x = element_text(size = 8))  # Rotate labels

hist_infection_length <- ggplot(master_data, aes(x = Length_of_Infection, fill = ARV.DRM)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7) +
  facet_wrap(~ ARV.DRM) +
  scale_fill_manual(values = c("Absent" = "blue", "Present" = "orange"))+
  labs(title = "Length of Infection by ARV.DRM Status (n=207)", x = "Length of Infection (years)", y = "Count") +
  theme_minimal()+
    theme(plot.title = element_text(size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title.x = element_text(size = 8))

# Function to create a donut chart with n values and labeled segments for Race
create_race_donut_chart <- function() {
  # Calculate counts, percentages, and update Race levels to include n
  df <- master_data %>%
    group_by(Race) %>%
    mutate(Race = paste0(Race, " (n = ", n(), ")")) %>%
    ungroup() %>%
    group_by(Race, ARV.DRM) %>%
    summarise(Count = n(), .groups = 'drop') %>%
    group_by(Race) %>%
    mutate(
      Percentage = Count / sum(Count) * 100,
      Label = paste0("n = ", Count)  # Create labels for each segment
    )

  # Generate the donut chart
  ggplot(df, aes(x = 2, y = Percentage, fill = ARV.DRM)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    xlim(0.5, 2.5) +  # Creates the hollow center for the donut chart
    scale_fill_manual(values = c("Absent" = "blue", "Present" = "orange")) +
    facet_wrap(~ Race, nrow = 3, ncol = 2) +
    # Use geom_label to add white boxes around text labels
    geom_label(aes(label = Label), position = position_stack(vjust = 0.6), size = 2, fill = "white",hjust=-.3) +
    labs(
      title = "ARV.DRM status within Race",
      x = NULL, y = NULL
    ) +
    theme_void() +
    theme(
      plot.title = element_text(size = 10),
      legend.position = "bottom",
      strip.text = element_text(size = 8),
    plot.margin = margin(10, 20, 10, 20)
    )
}

# Call the function to create the plot
race_plot <- create_race_donut_chart()

# Combine histograms into a single row
histograms_row <- hist_hiv_diagnosis + hist_infection_length + hist_age + race_plot + plot_layout(ncol = 4,widths = c(1,1,1,1.5))

# Combine all visualizations into one figure with adjusted heights
combined_plot <- histograms_row / pie_charts_grid +
  plot_layout(heights = c(1.5, 2)) +  # Adjust relative heights: 1 for histograms, 2 for pie charts
  plot_annotation(title = "Visual Analysis of ARV.DRM Status Across Variables")

# Save the final combined plot
ggsave("Final_Combined_ARV_DRM_Analysis_Updated.png", plot = combined_plot, width = 14, height = 10)

master_data <- master_data_restore 
```


# Check for colinearity

No evidence for colinearity between cannabis use and race

```{r}
table(master_data$Race, master_data$Cannabis)
chisq.test(table(master_data$Race, master_data$Cannabis))
fisher.test(table(master_data$Race, master_data$Cannabis))
```

# Check for colinearity with continuous variables

No evidence for co linearity between age at enrollment, length of infection, and cannabis use

```{r}
# Convert Cannabis from factor to numeric
master_data$Cannabis <- as.numeric(as.character(master_data$Cannabis))
# Run the correlation test
cor.test(master_data$Age_Enrollment, master_data$Cannabis)
cor.test(master_data$Length_of_Infection,master_data$Cannabis)
```

# Check for multicolinearity in model

No multicolinearity problems with demographic data, but we see the issue of colinearity between cannabis and poly_substance use

```{r}
library(car)
model_with_demo <- glm(AVDRM ~ Cannabis * PolySubstance + Amphetamine + Cocaine + Sedatives + Opiates + 
                       Race + Age_Enrollment,
                       data = master_data)
vif(model_with_demo)
```

# Stratify by poly substance use to address multicolinearity of cannabis and poly substance use

There aren't enough data points in the non poly substance use group to draw a conclusion from this approach (only 1 ARVDRM+/Cannabis+ data point in non poly substance use subset)

```{r}
# Subset data for non-poly_substance users
data_non_poly <- subset(data_factor, Poly_Substance == 0)
model_non_poly <- logistf(ARVDRM ~ Cannabis + Amphetamine + Cocaine + Sedatives + Opiates,
                      data = data_non_poly)

# Subset data for poly_substance users
data_poly <- subset(data_factor, Poly_Substance == 1)
model_poly <- logistf(ARVDRM ~ Cannabis + Amphetamine + Cocaine + Sedatives + Opiates,
                  data = data_poly)
```

```{r}
# Update values in data_poly
data_poly$Cannabis <- ifelse(data_poly$Cannabis == 1, "cannabis_pos (1)", "cannabis_neg (0)")
data_poly$ARVDRM <- ifelse(data_poly$ARVDRM == 1, "ARVDRM_pos (1)", "ARVDRM_neg (0)")

# Update values in data_non_poly
data_non_poly$Cannabis <- ifelse(data_non_poly$Cannabis == 1, "cannabis_pos (1)", "cannabis_neg (0)")
data_non_poly$ARVDRM <- ifelse(data_non_poly$ARVDRM == 1, "ARVDRM_pos (1)", "ARVDRM_neg (0)")
```

```{r}
library(ggplot2)
library(gridExtra)
library(grid)
library(gtable)
# Get the summaries of both models
summary_poly <- summary(model_poly)
summary_non_poly <- summary(model_non_poly)

# Extract Cannabis coefficient and p-value from summary_poly
# Coefficients and p-values are stored as named numeric vectors
coef_poly <- summary_poly$coefficients["Cannabis1"]
p_value_poly <- summary_poly$prob["Cannabis1"]

# Extract Cannabis coefficient and p-value from summary_non_poly
coef_non_poly <- summary_non_poly$coefficients["Cannabis1"]
p_value_non_poly <- summary_non_poly$prob["Cannabis1"]

# Format model formulas for display
formula_poly <- deparse(model_poly$call)
formula_non_poly <- deparse(model_non_poly$call)

# Format model summaries as text
model_summary_poly <- paste0(
  "Model (Poly): ", formula_poly, "\n",
  "Cannabis Coefficient: ", round(coef_poly, 3), "\n",
  "P-Value: ", round(p_value_poly, 5)
)

model_summary_non_poly <- paste0(
  "Model (Non-Poly): ", formula_non_poly, "\n",
  "Cannabis Coefficient: ", round(coef_non_poly, 3), "\n",
  "P-Value: ", round(p_value_non_poly, 5)
)

# Step 2: Create data tables for counts
df_table_poly <- as.data.frame.matrix(table(data_poly$Cannabis, data_poly$ARVDRM))
df_table_non_poly <- as.data.frame.matrix(table(data_non_poly$Cannabis, data_non_poly$ARVDRM))

# Modify the formulas to include line breaks
formula_poly <- "ARVDRM ~ Cannabis + Amphetamine\n+ Cocaine + Sedatives + Opiates"
formula_non_poly <- "ARVDRM ~ Cannabis + Amphetamine\n+ Cocaine + Sedatives + Opiates"

# Update the text for model summaries with the adjusted formulas
model_summary_poly <- textGrob(
  paste("Model (Poly):\n", formula_poly, "\n",
        "Cannabis Coefficient:", round(coef_poly, 3), "\n",
        "P-Value:", round(p_value_poly, 5)),
  gp = gpar(fontsize = 10), just = "left"
)

model_summary_non_poly <- textGrob(
  paste("Model (Non-Poly):\n", formula_non_poly, "\n",
        "Cannabis Coefficient:", round(coef_non_poly, 3), "\n",
        "P-Value:", round(p_value_non_poly, 5)),
  gp = gpar(fontsize = 10), just = "left"
)



# Step 4: Create bar plots
plot_poly <- ggplot(as.data.frame(table(data_poly$Cannabis, data_poly$ARVDRM)), 
                    aes(x = Var1, y = Freq, fill = as.factor(Var2))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Distribution in Poly-Substance Users",
       x = "Cannabis Status", y = "Count", fill = "ARVDRM") +
  theme_minimal()

plot_non_poly <- ggplot(as.data.frame(table(data_non_poly$Cannabis, data_non_poly$ARVDRM)), 
                        aes(x = Var1, y = Freq, fill = as.factor(Var2))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Distribution in Non Poly-Substance Users",
       x = "Cannabis Status", y = "Count", fill = "ARVDRM") +
  theme_minimal()

# Create the final plot
final_plot <- grid.arrange(
  arrangeGrob(tableGrob(df_table_poly), model_summary_poly, ncol = 1),
  plot_poly,
  arrangeGrob(tableGrob(df_table_non_poly), model_summary_non_poly, ncol = 1),
  plot_non_poly,
  ncol = 2,
  top = "Summary of Cannabis and ARV-DRM by Poly-Substance Use Status"
)

final_plot
```

### Include demographic covariates in best model

```{r}
# Recode Race to be "White" or "Non-White"
data_factor_demo$Race <- ifelse(data_factor_demo$Race == "White", "White", "Non-White")

# Convert Race to a factor
data_factor_demo$Race <- as.factor(data_factor_demo$Race)
```

### Get the odds ratio for the best model

```{r}
data_factor_
best_model <- glm(ARVDRM ~ Cannabis +  Poly_Substance + Amphetamine + Cocaine + Sedatives + Opiates +
                  Race + Length_of_Infection + Age_Enrollment,
                  data = data_factor_demo, family = binomial())

# Calculate odds ratios and confidence intervals
best_model_coefs <- coef(best_model)
best_model_or <- exp(best_model_coefs)  # Odds ratios
conf_int <- exp(confint(best_model))    # Confidence intervals

# Extract p-values from the model summary
best_model_summary <- summary(best_model)
p_values <- coef(best_model_summary)[, "Pr(>|z|)"]  # Extract p-values

# Combine odds ratios, confidence intervals, and p-values into a data frame
odds_ratio_df <- data.frame(
  Predictor = names(best_model_or),
  Odds_Ratio = round(best_model_or, 3),
  CI_95_Lower = round(conf_int[, 1], 3),
  CI_95_Upper = round(conf_int[, 2], 3),
  P_Value = round(p_values, 5)
)

# View the table with odds ratios, confidence intervals, and p-values
print(odds_ratio_df)

# Save the table to a CSV file
write.csv(odds_ratio_df, "results/suppl.Table.csv", row.names = FALSE)
```

See if using firth logistic regression for rare events changes results of best model

```{r}
# Load the logistf package
library(logistf)

# Perform Firth logistic regression
firth_model <- logistf(ARVDRM ~ Cannabis * Poly_Substance + Amphetamine + Cocaine + Sedatives + Opiates,
                  data = data_factor)

# Summarize the results
summary(firth_model)
```

```{r}
# Calculate odds ratios and confidence intervals
best_model_or <- exp(coef(firth_model))  # Odds ratios
conf_int <- exp(confint(firth_model))    # Confidence intervals

# Extract p-values from the model summary
best_model_summary <- summary(firth_model)
p_values <- best_model_summary$prob  # Extract p-values from the summary

# Combine odds ratios, confidence intervals, and p-values into a data frame
odds_ratio_df <- data.frame(
  Predictor = names(best_model_or),
  Odds_Ratio = round(best_model_or, 3),
  CI_95_Lower = round(conf_int[, 1], 3),
  CI_95_Upper = round(conf_int[, 2], 3),
  P_Value = round(p_values, 5)
)

# View the table with odds ratios, confidence intervals, and p-values
print(odds_ratio_df)

```
